{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-03 paper.tex
Sec 3 edits

\tableofcontents

@@

\subsection{Properties of {\tt FK} tables}
Some discussion here of the expected performance
differences. Including precomputed sums over scale, perturbative
order, no need to perform PDF evolution on the fly, smaller initial
scale PDF basis, simple convolution structure that can be accellerated
with SIMD/OpenMP.  Explain reduced flexibility w.r.t $\alpha_S$
variations and scales, discuss increased storage space requirement but
also reduced memory footprint.  Might want to mention possibility for
generating 'fast' grids as the interpolation accuracy is decoupled.

@@

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
For a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross-sections~\cite{ATLAS:2012aa,ATLAS:2011xha,TheATLAScollaboration:2013dja,Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross-sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are provided for a wide range of kinematics, for a total of 52 source
{\tt APPLgrid} files roughly corresponding to the full collider
dataset of the forthcoming NNPDF3.1 PDF determination.

In Fig.~\ref{fig:timings} we compare the convolution time per point
between {\tt APFELgrid} and {\tt APPLgrid} for all the 52
processes. We show timings for the {\tt APFELgrid} library in four
different configurations: AVX-OpenMP 2x (2 CPUs), AVX SSE3, and
standard double precision dot product convolution. When comparing the
results we observe a great performance improvement for convolution
time per point when using {\tt APFELgrid}. In particular, when
comparing {\tt APFELgrid} AVX to {\tt APPLgrid} timings we obtain a
minimum of 10x improvement for electroweak vector boson production and
a maximum of 2000x for inclusive jet data. These results confirm the
great usefulness of {\tt APFELgrid} in terms of performance for fits
of PDFs.

In Fig.~\ref{fig:performance} we provide information about the {\tt
  APFELgrid} table generation, grid size and memory usage. In terms of
generation time per point, we observe timings from few milliseconds to
3.5 minutes per point. Such difference is almost due to the density of
weights contained in the input {\tt APPLgrid} which requires a larger
number of combination operations. When considering the grid size we
confirm that in several processes the {\tt APPLgrid} files are smaller
than {\tt APFELgrid} ones as expected. This feature is justified by
the lack of compression of {\tt APFELgrid} files. Finally, when
measuring the resident set size memory used by the convolution
programs, the amount of memory required by {\tt APFELgrid} is smaller
than {\tt APPLgrid} for all processes considered here by at least
75\%.

In conclusion, the great performance benefits obtained with {\tt
  APFELgrid} endorse the importance of this tool not only in PDF fits
but also in any application requiring a large number of convolutions.

>>>>

%\tableofcontents

@@

\subsection{Properties of {\tt FK} tables}
Some discussion here of the expected performance
differences. Including precomputed sums over scale, perturbative
order, no need to perform PDF evolution on the fly, smaller initial
scale PDF basis, simple convolution structure that can be accelerated
with SIMD/OpenMP.  Explain reduced flexibility w.r.t $\alpha_S$
variations and scales, discuss increased storage space requirement but
also reduced memory footprint.  Might want to mention possibility for
generating 'fast' grids as the interpolation accuracy is decoupled.

@@

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt FK}).
%
To provide a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross-sections~\cite{ATLAS:2012aa,ATLAS:2011xha,TheATLAScollaboration:2013dja,Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross-sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are performed over a wide range of kinematics, for a total of 52 source
{\tt APPLgrid} files corresponding to the majority of available LHC and Tevatron datasets of applicability to PDF determination.
While the source {\tt APPLgrid} files have varying grid densities in $x$ and $Q^2$, for the purposes of comparison the corresponding {\tt FK} tables are produced
consistently with 30 points in $x$, and at an initial scale below the charm threshold, therefore with seven active partonic species.

In Fig.~\ref{fig:timings} we compare the average time taken per datapoint
for the {\tt FK} and {\tt APPLgrid} calculations, for all of the 52
tables. We show timings for the {\tt FK} calculation in four
different configurations: AVX-OpenMP 2x (2 CPU cores), AVX, SSE3 and
standard double precision dot product. Due to the inherent structural differences between the {\tt FK} and {\tt APPLgrid} procedures,
results from the {\tt FK} calculation are systematically faster than those from {\tt APPLgrids}.
In particular, when comparing {\tt FK} AVX to {\tt APPLgrid} timings we obtain minimally a factor
of ten improvement in speed for electroweak vector boson production and
a maximum factor of 2000 improvement in predictions for inclusive jet data. Across all processes
and kinematic regions we observe significant performance improvements from using the {\tt FK} calculation
even in the maximally compatible, non-SIMD case.

While sheer computational speed is typically the primary consideration when computing observables
in a PDF fit, other factors such as table size in the filesystem and memory, along with the computational cost of
pre-computing {\tt FK} tables must be considered. Indeed, the computation of the {\tt FK} table in Eqn.~\ref{eq:FKTable}
requires a great deal of operations which can be time consuming, particuarly in the case of source {\tt APPLgrids} with very high
interpolation precision.

In Fig.~\ref{fig:performance} we examine the {\tt FK} table generation time with {\tt APFELgrid}, {\tt FK} table file size and memory usage of the {\tt FK} tables arising from the same source {\tt APPLgrids} as discussed in Fig.~\ref{fig:timings}. 
When examining the table generation time per point, we observe timings from a few milliseconds to
3.5 minutes per point, with differences arising from the varying grid densities used in the input {\tt APPLgrid} files. 
In terms of the grid size on disc, {\tt FK} tables are typically larger than their corresponding {\tt APPLgrids}, primarily as the {\tt FK} file format is encoding in plain text for compatibility whereas {\tt APPLgrids} are expressed in binary as {\tt ROOT} files. However when
measuring the in-memory resident set size used by the both convolution procedures, the amount of memory required by {\tt FK} tables is systematically less than {\tt APPLgrid} for all processes considered here by at least 75\%.

%In conclusion, the great performance benefits obtained with {\tt
%APFELgrid} endorse the importance of this tool not only in PDF fits
%but also in any application requiring a large number of convolutions.
